Raw Loans Analysis
========================================================


### Load libraries


```{r}
library(ggplot2)
library(lattice)
library(plyr)
library(reshape2)
library(RColorBrewer)
```

## Processing

Download the data, read the data in and save the raw data as an RDA file

```{r loadData, cache=TRUE}
download.file('https://spark-public.s3.amazonaws.com/dataanalysis/loansData.csv', destfile = './data/loansData.csv')
downloadDate <- date()
downloadDate
loansRaw <- read.csv('./data/loansData.csv')
save(loansRaw, downloadDate , file='./data/loans.rda')
```


### Look at the data

```{r, dependson='loadData'}
names(loansRaw)
str(loansRaw)
head(loansRaw)
```


### Find out about missing values

#### Explicit NAs

```{r, dependson='loadData'}
sum(is.na(loansRaw))
dim(loansRaw)
sum(complete.cases(loansRaw))
loansRaw[!complete.cases(loansRaw),]
```

#### Implicit NAs
```{r, dependson='loadData'}
table(loansRaw$Employment.Length)
```


### Converting character vectors into numerical vectors

```{r processData, dependson='loadData'}
loansProcessed <- loansRaw
interestRate <- as.numeric(sapply(loansRaw$Interest.Rate, function(x) sub('%','',x)))
debtToIncomeRatio <- as.numeric(sapply(loansRaw$Debt.To.Income.Ratio, function(x) sub('%','',x)))
loansProcessed$Interest.Rate <- interestRate
loansProcessed$Debt.To.Income.Ratio <- debtToIncomeRatio
class(loansProcessed$Interest.Rate)
loansProcessed$Interest.Rate[1:10]
class(loansProcessed$Debt.To.Income.Ratio)
loansProcessed$Debt.To.Income.Ratio[1:10]
```

### Transforming 'n/a' values into NAs, exclude observations with NAs

```{r}
loansProcessed$Employment.Length <- sub('n/a', NA, loansProcessed$Employment.Length)
table(loansProcessed$Employment.Length, useNA='ifany')
loansProcessed <- loansProcessed[complete.cases(loansProcessed),]
dim(loansProcessed)

```

### Transforming Employment.Length

```{r}
empl <- sub('< 1 year',0,loansProcessed$Employment.Length)
empl <- sub('10\\+ years',11,empl)
regpos <- regexec('([0-9]{1,2})(.)*', empl)
empl <- sapply(regmatches(empl,regpos), function(x)x[2])
empl <- sub('0','0.5', empl)
empl <- as.numeric(empl)

loansProcessed$Employment.Length <- empl

```

### Exluding amounts funded <= 0 rename Loan.Length values

```{r}
loansProcessed <- subset(loansProcessed,Amount.Funded.By.Investors>0)
loansProcessed$Loan.Length <-  as.factor(sub(' months','',loansProcessed$Loan.Length))

```


## Exploratory Analysis

### singular variable analysis

#### Explore interest rate

```{r, dependson='processData'}
boxplot(loansProcessed$Interest.Rate)
hist(loansProcessed$Interest.Rate)
ggplot(loansProcessed,aes(x=Interest.Rate)) + geom_density()

boxplot(Interest.Rate ~ FICO.Range,loansProcessed)#
boxplot(Interest.Rate ~ Home.Ownership,loansProcessed)
boxplot(Interest.Rate ~ Employment.Length,loansProcessed)
boxplot(Interest.Rate ~ Loan.Purpose,loansProcessed)
boxplot(Interest.Rate ~ Loan.Length,loansProcessed)#
boxplot(Interest.Rate ~ State,loansProcessed)

```

#### Explore FICO range

```{r, dependson='processData'}
levels(loansProcessed$FICO.Range)
table(loansProcessed$FICO.Range)
barplot(table(loansProcessed$FICO.Range))
```

#### Explore amount requested

```{r, dependson='processData'}
boxplot(loansProcessed$Amount.Requested)
hist(loansProcessed$Amount.Requested)
ggplot(loansProcessed,aes(x=Amount.Requested)) + geom_density()
```


#### Explore amount founded by investors

```{r, dependson='processData'}
boxplot(loansProcessed$Amount.Funded.By.Investors)
hist(loansProcessed$Amount.Funded.By.Investors)
ggplot(loansProcessed,aes(x=Amount.Funded.By.Investors)) + geom_density()

```


#### Explore debt-to-income ratio

A debt-to-income ratio (often abbreviated DTI) is the percentage of a consumer's monthly gross income that goes toward paying debts

```{r, dependson='processData'}
boxplot(loansProcessed$Debt.To.Income.Ratio)
hist(loansProcessed$Debt.To.Income.Ratio)
ggplot(loansProcessed,aes(x=Debt.To.Income.Ratio)) + geom_density()

```


#### Explore loan purpose

```{r, dependson='processData'}
table(loansProcessed$Loan.Purpose)
barplot(table(loansProcessed$Loan.Purpose))

```

#### Explore monthly income

```{r, dependson='processData'}

boxplot(loansProcessed$Monthly.Income)
hist(loansProcessed$Monthly.Income,breaks=100)
```

##### Removing 3 outliers

```{r, dependson='processData'}
tmp_month <- loansProcessed$Monthly.Income[loansProcessed$Monthly.Income < 30000]
boxplot(tmp_month)
hist(tmp_month,breaks=100)
```


#### Explore inquiries in last 6 months

```{r, dependson='processData'}
barplot(table(loansProcessed$Inquiries.in.the.Last.6.Months))

```

#### Explore Revolving credit balance

```{r, dependson='processData'}
hist(loansProcessed$Revolving.CREDIT.Balance)

```





### make some scatterplots

#### look at the correlation first


```{r}
sapply(loansProcessed,function(x)cor(as.numeric(x),loansProcessed$Interest.Rate))
cor(loansProcessed$Interest.Rate,as.numeric(as.factor(loansProcessed$Employment.Length)))
```

#### amount requested vs. interest rate

```{r, dependson='processData'}
ggplot(loansProcessed,aes(x=Amount.Requested, y=Interest.Rate)) + geom_point(size=3) 

```

#### amount funded vs. interest rate

```{r, dependson='processData'}
ggplot(loansProcessed,aes(x=Amount.Funded.By.Investors,y=Interest.Rate, colour=as.factor(Employment.Length))) + geom_point()
```
#### amount requested vs amount funded

```{r, dependson='processData'}
ggplot(loansProcessed,aes(x=Amount.Requested, y=Amount.Funded.By.Investors)) + geom_point()

```

#### debt-to-income vs. interest rate

```{r, dependson='processData'}
ggplot(loansProcessed,aes(x=Debt.To.Income.Ratio, y=Interest.Rate)) + geom_point(size=3,aes(color=Loan.Length))

```

#### FICO range vs. interest rate

```{r, dependson='processData'}
ggplot(loansProcessed,aes(x=as.numeric(FICO.Range), y=Interest.Rate)) + geom_point(size=3)  + stat_smooth(method=lm)

ggplot(loansProcessed,aes(x=FICO.Range,y=Interest.Rate,colour=as.factor(Employment.Length))) + geom_point()
ggplot(loansProcessed,aes(x=FICO.Range,y=Interest.Rate,colour=Loan.Length)) + geom_point()

```

#### Revolving credit balance vs. interest rate

```{r,dependson='processData'}
plot(loansProcessed$Revolving.CREDIT.Balance,loansProcessed$Interest.Rate)
rev <- subset(loansProcessed,Revolving.CREDIT.Balance<150000)
plot(rev$Revolving.CREDIT.Balance,rev$Interest.Rate)

plot(loansProcessed$Open.CREDIT.Lines,loansProcessed$Interest.Rate)

```

#### inquiries in last 6 months vs interest rate

```{r, dependson='processData'}
ggplot(loansProcessed,aes(x=Inquiries.in.the.Last.6.Months,y=Interest.Rate)) + geom_point()

```

### Look at dependencies between Interest Rate and other quantitative variables for each FICO range


```{r}
xyplot(Interest.Rate ~ Amount.Requested | FICO.Range,data=loansProcessed)

xyplot(Interest.Rate ~ Debt.To.Income.Ratio | FICO.Range,data=loansProcessed)

xyplot(Interest.Rate ~ Monthly.Income | FICO.Range,data=loansProcessed)

xyplot(Interest.Rate ~ Monthly.Income | FICO.Range,data=subset(loansProcessed,Monthly.Income < 15000))

xyplot(Interest.Rate ~ Open.CREDIT.Lines | FICO.Range,data=loansProcessed)

xyplot(Interest.Rate ~ Revolving.CREDIT.Balance | FICO.Range,data=loansProcessed)

xyplot(Interest.Rate ~ Revolving.CREDIT.Balance | FICO.Range,data=subset(loansProcessed,Revolving.CREDIT.Balance < 14000))

xyplot(Interest.Rate ~ Revolving.CREDIT.Balance | FICO.Range,data=loansProcessed)

xyplot(Interest.Rate ~ Inquiries.in.the.Last.6.Months | FICO.Range,data=loansProcessed)

xyplot(Interest.Rate ~ Employment.Length | FICO.Range,data=loansProcessed)

```



### Examining FICO subsets

#### subsetting loansProcessed

#### Range 670-674

```{r subset ,dependson='processData'}
sort(table(loansProcessed$FICO.Range))
loansFicoSubset1 <- subset(loansProcessed,FICO.Range == '670-674')
nrow(loansFicoSubset1)

empl <- sub('< 1 year',0,loansFicoSubset1$Employment.Length)
empl <- sub('10\\+ years',11,empl)
regpos <- regexec('([0-9]{1,2})(.)*',empl)
empl <- sapply(regmatches(empl,regpos),function(x)x[2])
empl <- sub('0','0.5',empl)
empl <- as.numeric(empl)

loansFicoSubset1$Employment.Length <- empl


```

#### make some histograms for subset

```{r,dependson='subset'}
boxplot(Interest.Rate ~ FICO.Range,loansFicoSubset1)
boxplot(Interest.Rate ~ Home.Ownership,loansFicoSubset1)
boxplot(Interest.Rate ~ Employment.Length,loansFicoSubset1)
boxplot(Interest.Rate ~ Loan.Purpose,loansFicoSubset1)
boxplot(Interest.Rate ~ Loan.Length,loansFicoSubset1)#
boxplot(Interest.Rate ~ State,loansFicoSubset1)
boxplot(Interest.Rate ~ as.factor(Inquiries.in.the.Last.6.Months),loansFicoSubset1)
boxplot(Interest.Rate ~ as.factor(Employment.Length ),loansFicoSubset1)
```

#### make some scatterplots

```{r,dependson='processData'}
plot(loansFicoSubset1$Amount.Funded.By.Investors,loansFicoSubset1$Interest.Rate,col=loansFicoSubset1$Loan.Length)

plot(loansFicoSubset1$Revolving.CREDIT.Balance,loansFicoSubset1$Interest.Rate,col=loansFicoSubset1$Loan.Length)

plot(loansFicoSubset1$Debt.To.Income.Ratio,loansFicoSubset1$Interest.Rate)
```

## Modeling

### interest rate ~ Fico(overall)

```{r,dependson='processData'}
lm1 <- lm(Interest.Rate ~ FICO.Range,loansProcessed)
summary(lm1)
par(mfrow = c(1, 2))
plot(as.numeric(loansProcessed$FICO.Range), lm1$residuals, pch = 19)
plot(lm1$fitted, lm1$residuals, pch = 19)

```


#### Interest.Rate ~ FICO.Range + Amount.Requested

```{r,dependson='processData'}
lm2 <- lm(Interest.Rate ~ FICO.Range + Amount.Requested,loansProcessed)
#summary(lm2)
par(mfrow = c(1, 2))
plot(loansProcessed$Amount.Requested, lm2$residuals, pch = 19)
plot(lm2$fitted, lm2$residuals, pch = 19)

```

#### Interest.Rate ~ FICO.Range + Amount.Requested (Colored by Loan.Length, State, Home Ownership, Employment Length)

```{r,dependson='processData'}

lm2 <- lm(Interest.Rate ~ FICO.Range + Amount.Requested,loansProcessed)
par(mfrow = c(2, 4))
plot(loansProcessed$Amount.Requested, lm2$residuals, pch = 19,col=loansProcessed$Loan.Length)
plot(lm2$fitted, lm2$residuals, pch = 19,col=loansProcessed$Loan.Length, main='loan length')

plot(loansProcessed$Amount.Requested, lm2$residuals, pch = 19,col=loansProcessed$State)
plot(lm2$fitted, lm2$residuals, pch = 19,col=loansProcessed$State,main='state')

plot(loansProcessed$Amount.Requested, lm2$residuals, pch = 19,col=loansProcessed$Home.Ownership)
plot(lm2$fitted, lm2$residuals, pch = 19,col=loansProcessed$Home.Ownership,main='home ownership')

plot(loansProcessed$Amount.Requested, lm2$residuals, pch = 19,col=loansProcessed$Employment.Length)
plot(lm2$fitted, lm2$residuals, pch = 19,col=loansProcessed$Employment.Length, main='employment length')

```



#### Interest.Rate ~ FICO.Range + Amount.Requested + Loan Length

```{r,dependson='processData'}
lm3 <- lm(Interest.Rate ~ FICO.Range + Amount.Requested + Loan.Length,loansProcessed)
#summary(lm2)
par(mfrow = c(1, 2))
plot(loansProcessed$Amount.Requested, lm3$residuals, pch = 19)
plot(lm3$fitted, lm3$residuals, pch = 19)

```

#### Coloring lm3

```{r, dependson='processData'}
par(mfrow = c(2, 4))
plot(loansProcessed$Amount.Requested, lm3$residuals, pch = 19,col=loansProcessed$Loan.Length)
plot(lm3$fitted, lm3$residuals, pch = 19,col=loansProcessed$Loan.Length,main='loan length')

plot(loansProcessed$Amount.Funded.By.Investors, lm3$residuals, pch = 19,col=loansProcessed$State)
plot(lm3$fitted, lm3$residuals, pch = 19,col=loansProcessed$State, main='State')

plot(loansProcessed$Amount.Requested, lm3$residuals, pch = 19,col=loansProcessed$Home.Ownership)
plot(lm3$fitted, lm3$residuals, pch = 19,col=loansProcessed$Home.Ownership,main='home ownership')

plot(loansProcessed$Amount.Requested, lm3$residuals, pch = 19,col=loansProcessed$Employment.Length)
plot(lm3$fitted, lm3$residuals, pch = 19,col=loansProcessed$Employment.Length, main='employment length')

```

#### Interest.Rate ~ FICO.Range + Loan Length * Amount.Requested + Loan Length 

```{r,dependson='processData'}
lm4 <- lm(Interest.Rate ~ FICO.Range + Loan.Length + Amount.Requested*Loan.Length, loansProcessed)

par(mfrow = c(1, 2))
plot(loansProcessed$Amount.Requested, lm4$residuals, pch = 19)
plot(lm4$fitted, lm4$residuals, pch = 19)

```




#### Plot the residuals versus the observed and fitted of all prior models here


```{r, dependson='processData'}
par(mfrow=c(2,3))
plot(lm1$residuals, pch = 19,col=loansProcessed$Loan.Length)
plot(lm1$fitted, lm1$residuals, pch = 19,col=loansProcessed$Loan.Length)

plot(loansProcessed$Amount.Requested, lm2$residuals, pch = 19,col=loansProcessed$Loan.Length)
plot(lm2$fitted, lm2$residuals, pch = 19,col=loansProcessed$Loan.Length)

plot(loansProcessed$Amount.Requested, lm3$residuals, pch = 19,col=loansProcessed$Loan.Length)
plot(lm3$fitted, lm3$residuals, pch = 19,col=loansProcessed$Loan.Length)
```

#### prior vs interaction

```{r, dependson='processData'}
par(mfrow=c(1,2))

plot(loansProcessed$Amount.Requested, lm3$residuals, pch = 19,main='lm3')
plot(loansProcessed$Amount.Requested, lm4$residuals, pch = 19)

plot(lm3$fitted, lm3$residuals, pch = 19,main='lm3')
plot(lm4$fitted, lm4$residuals, pch = 19)

```

#### Model comparison (lm1 - lm4)

```{r}
summary(lm1)
summary(lm2)
summary(lm3)
summary(lm4)

qqnorm(lm1$residuals)
qqline(lm1$residuals)

qqnorm(lm2$residuals)
qqline(lm2$residuals)

qqnorm(lm3$residuals)
qqline(lm3$residuals)

confint(lm3)

amount = loansProcessed$Amount.Requested/1000

lm5 <-lm(Interest.Rate ~ as.numeric(FICO.Range) + amount + Loan.Length,loansProcessed)
confint(lm5)

```


### Final Graphics

```{r,dependson='processData'}
sort(table(loansProcessed$FICO.Range))
lp.subset <- subset(loansProcessed,FICO.Range=='670-674'|FICO.Range=='675-679'|FICO.Range=='680-684')
lp.subset


g1 <- ggplot(lp.subset,aes(y=Interest.Rate,x=Amount.Requested)) + geom_point(size=2) + facet_grid(FICO.Range~.) + stat_smooth(method=lm,se=FALSE) + xlab('Amount Requested (dollars)') + ylab('Interest Rate (percent)') + theme(axis.title.x=element_text(size=15),axis.title.y=element_text(size=15),strip.text=element_text(size=15),axis.text.y=element_text(size=12),axis.text.x=element_text(size=15)) 
g1

g2 <- ggplot(,aes(x=loansProcessed$Amount.Requested,y=lm2$residuals)) + geom_point(size=2.5,aes(color=loansProcessed$Loan.Length)) + scale_colour_brewer(palette='Set1') + labs(color='Loan length\n(months)') + xlab('Amount Requested (dollars)') + ylab('No Adjustment Residuals') + theme(axis.title.x=element_text(size=15),axis.title.y=element_text(size=15),legend.title=element_text(size=11),axis.text=element_text(size=15)) + guides(color=FALSE)
g2

g4 <- ggplot(,aes(x=loansProcessed$Amount.Requested,y=lm4$residuals)) + geom_point(size=2.5,aes(color=loansProcessed$Loan.Length)) + scale_colour_brewer(palette='Set1') + labs(color='Loan length\n(months)') + xlab('Amount Requested (dollars)') + ylab('Full Model Residulas') + theme(axis.title.x=element_text(size=15),axis.title.y=element_text(size=15),legend.title=element_text(size=11),axis.text=element_text(size=15),legend.position=c(0.93,0.1))
g4



```



